---
title: "Final Project - DATA 612"
author: "Bruno de Melo and Leland Randles"
date: "July 16th, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(tidyverse)) install.packages("tidyverse",repos = "http://cran.us.r-project.org")
if(!require(recommenderlab)) install.packages("recommenderlab",repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("tidyverse",repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra",repos = "http://cran.us.r-project.org")
if(!require(sparklyr)) install.packages("sparklyr",repos = "http://cran.us.r-project.org")
if(!require(NCmisc)) install.packages("NCmisc",repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table",repos = "http://cran.us.r-project.org")
library("tidyverse")
library("recommenderlab")
library("kableExtra")
library("sparklyr")
library("NCmisc")
library("data.table")
```  
  
## Loading Datasets

Per our final project proposal, there are two datasets we are using, which came from the web site https://myanimelist.net/, which is the world's largest anime and manga community.  

The datsets are available at kaggle: https://www.kaggle.com/CooperUnion/anime-recommendations-database.

### Content for Each Dataset  
  
* Anime.csv  
     
    + anime_id - myanimelist.net's unique id identifying an anime.     
    + name - full name of anime.    
    + genre - comma separated list of genres for this anime.     
    + type - movie, TV, OVA, etc.      
    + episodes - how many episodes in this show. (1 if movie).     
    + rating - average rating out of 10 for this anime.      
    + members - number of community members that are in this anime's "group".    

* Rating.csv              
     
    + user_id - non identifiable randomly generated user id.      
    + anime_id - the anime that this user has rated.     
    + rating - rating out of 10 this user has assigned (-1 if the user watched it but didn't assign a rating).  
  
We will use the ratings for collaborative filtering and genre plus type for content-based recommendation. Below, we load the datasets:   
  
```{r load, warning=FALSE, message=FALSE}
set.seed(1234)
#loading dataset
anime_raw <- read_csv("https://raw.githubusercontent.com/Randles-CUNY/DATA612/master/Final_Project/datasets_571_1094_anime.csv") %>% data.frame()  
anime_raw$name = str_replace_all(anime_raw$name,"&#039;","'")  
anime_raw$name = str_replace_all(anime_raw$name,"&quot;","")
anime_raw$name = str_replace_all(anime_raw$name,".hack//","") 
anime_raw$name = str_replace_all(anime_raw$name,"_","")
anime_raw$name = str_replace_all(anime_raw$name,"&amp;","&")
#useful data frames
anime_name <- select(anime_raw, c(1, 2))
anime_genre <- select(anime_raw, c(1, 3, 4))
anime_members <- select(anime_raw, c(1, 7))
anime_comp <- anime_raw
#loading dataset
ratings_raw1 <- read_csv("https://raw.githubusercontent.com/bsvmelo/DATA612/master/Final_Project/rating_partaa.csv") %>% data.frame()
ratings_raw2 <- read_csv("https://raw.githubusercontent.com/bsvmelo/DATA612/master/Final_Project/rating_partab.csv", col_names = FALSE) %>% data.frame()
#merging both datasets
colnames(ratings_raw2) <- c(colnames(ratings_raw1))
ratings_raw <- rbind(ratings_raw1, ratings_raw2)
# -1 was originally assigned if the user watched it but didn't assign a rating. We will replace it by NA.
ratings_raw$rating[(ratings_raw$rating) == -1 ] <- NA
```  
  
## Data exploration  
  
### By type  
  
Based on the plot below, we decided exclude Music and NA from our recommender system dataset. We included TV, OVA, Movie, Special and ONA.  
  
```{r , warning=FALSE, message = FALSE }
# Type
a_type <- anime_comp %>% group_by(type) %>% summarise(Count=n())
ggplot(a_type, mapping = aes(x=reorder(type, -Count), y=Count)) + geom_col() + labs(x = "Anime Types")
```

### Relationship Between Number of Members and Ratings 
  
Per the data dictionary provided earlier, the `anime.csv` dataset includes two data elements of interest:  "members", which tells you how many community members joined the "group" associated with an anime show, and "rating" which is the average rating (out of 10) given by the anime group. Below, We check the relationship between the number of members and the ratings to check for biases.    
  
We used the log of the number of members in order to account for its variability (ranges from 5 to 1 million plus).  
  
The chart below indicates that there's a positive relationship between value of `members` and `rating`. We can also see in the rectangular area that there are less popular groups which have higher ratings, perhaps reflecting great anime shows which are more obscure. We may use this fact to enrich our recommendation list.   
  
```{r ,message=FALSE}
anime_comp <- drop_na(anime_comp)
ggplot(anime_comp, mapping = aes(x = log(members), y=rating)) + geom_point(alpha=1/15) + geom_smooth() + annotate("rect", xmin=c(3.3), xmax=c(4.7), ymin=c(7.3), ymax=c(10), alpha=0.2, color = "blue", fill = "orange")
```
  
## Content-Based Recommendation  
  
First, we built a content-based recommendation system based on genre similarity and popularity. The approach was to take the data element `genre` (which can contain multiple tagged genres) and build recommendations accordingly.  
  
### Recommendations Based on Genre Similarity  
  
First, we filtered the dataset by genre, getting rid of NAs values. Then, we create a binary matrix of animes by genres. Lastly, we calculate a similarity matrix.  
  
The utimate goal is to randomly select an anime, and display a top N recommendation list based on similarity between genres.  
  
```{r genre}
# subsetting for type and NAs
anime_genre <- subset(anime_genre, type %in% c("TV", "OVA", "Movie", "Special", "ONA"), -type)
anime_genre <- drop_na(anime_genre) 
anime_genre_ids <- anime_genre$anime_id
# Genre dataset conversion to realRatingsMatrix
# splitting genre column
genres1 <- as.data.frame(tstrsplit(anime_genre[,2], '[,]', type.convert = FALSE), stringsAsFactors = FALSE)
colnames(genres1) <- c(1:ncol(genres1))
# grabbing unique genres
genre_vec <- (as.vector(trimws(unlist(genres1))))
uniq_genre <- unique(genre_vec)
uniq_genre <- uniq_genre[!is.na(uniq_genre)]
#creating matrix anime_id and genres
anime_genre_mtx <- cbind(anime_id = anime_genre$anime_id, genres1)
#creating binarized matrix anime_id and genre
genre_mtx <- matrix(NA, nrow(anime_genre), length(uniq_genre) + 1)
genre_mtx[,1] <- anime_genre_mtx$anime_id
colnames(genre_mtx) <- c("anime_id", uniq_genre)
for (i in 1:nrow(anime_genre_mtx)) {
 for (j in 2:ncol(anime_genre_mtx)) {
  mat_col = which(uniq_genre == trimws(anime_genre_mtx[i, j])) + 1
  genre_mtx[i, mat_col] <- 1
 }
}
genre_b <- genre_mtx
genre_b[, 2:ncol(genre_mtx)][is.na(genre_b[, 2:ncol(genre_mtx)])] <- 0
genre_b[, 2:ncol(genre_mtx)][genre_b[, 2:ncol(genre_mtx)] > 0] <- 1
genre_b <- genre_b[, 2:ncol(genre_mtx)]
# calculate similarity matrix - non normalized
genre_bm <- as(genre_b, "binaryRatingMatrix")
genre_sim <- similarity(genre_bm, method = "Jaccard", which = "users")
# visualization - 20 animes
genre_v <- as(genre_sim, "matrix")
colnames(genre_v) <- anime_genre$anime_id
rownames(genre_v) <- anime_genre$anime_id
image(genre_v[1:20, 1:20], main = "Similarity Matrix Between 20 Animes")
```  
  
#### Top N Recommendations by Genre Similarity  
  
Below is the implementation of the recommendation list based on the similarity matrix we calculated above. One anime show is randomly sampled and then the N most similar animes to the one sampled are displayed.

```{r genre rec}
# set number of recommendations
n_recommended <- 10
# randomly select an anime_id and get corresponding name 
a_id <- sample(anime_genre$anime_id, 1)
a_name <- anime_name[anime_name$anime_id == a_id, ]$name
# get recommendations
recs <- sort(genre_v[as.character(a_id),], decreasing = TRUE)[1:n_recommended]
# get IDs
recs_id <- as.numeric(names(recs))
# create list 
recs_names <- anime_name[anime_name$anime_id %in% recs_id,]$name
# create table
header <- sprintf("Animes Similar to %s", a_name)
# display the list of similar animes
kable(recs_names, col.names = header) %>% kable_styling()
```
  
### Recommendations Based on Popularity   
  
We measured popularity by the number of times an anime has been viewed, based on the `ratings.csv` dataset. The ratings dataset is a typical item-user matrix. A frequency term was created using the function `colCounts` from the `recommenderLab` package.  
  
First, we conformed the rating dataset to the dataset used for Genre Similarity. Then, we created a viewer frequency or popularity dataset. We then transposed the binary matrix of animes by genres we'd created.  
  
In order to incorporate the popularity dataset into the binary matrix, our approach was to divide the number of views by the number of genres for the anime show. To avoid penalizing anime shows that have many genre classifications, the result was weighted by the square root of the total numbers of genres.  
  
The ultimate goal was to randomly select a genre and display a top N recommendation list based on the popularity index.  
  
```{r}
# creating matrix genre-anime
# ratings matrix
# subsetting to match same user id used in the matrix above
rating_raw <- ratings_raw[ratings_raw$user_id %in% anime_genre_ids, ]
rating_mtx <- as(rating_raw, "binaryRatingMatrix")
user_mtx <- t(as(rating_mtx, "matrix"))
# counting how many times an anime has been seen
genre_count <- data.frame(anime_id = names(colCounts(rating_mtx)), views = colCounts(rating_mtx))
# creating matrix anime-genre
rownames(genre_b) <- anime_genre$anime_id
genre_t <- t(genre_b)
colnames(genre_t) <- anime_genre$anime_id
genre_t <- genre_t[, colnames(genre_t) %in% c(as.character(genre_count$anime_id))]
#insert number of views in each column, weighted by square root of total numbers of genres
for (i in 1:ncol(genre_t)) {
  mat_col = which(colnames(genre_t)[i] == as.character(genre_count$anime_id))
  s <- genre_count$views[mat_col]/sqrt(sum(genre_t[, i]))
  genre_t[, i] <- genre_t[, i] * s
}
```
 
#### Top N Recommendations by Popularity
  
Below is the implementation of the recommender list based on the popularity matrix we'd calculated. To demonstrate, a genre is randomly sampled and the N most popular animes related to that genre are displayed.  
  
```{r genre rec}
# set number of recommendations
n_recommended <- 10
# randomly select a anime_id and get corresponding name 
s_genre <- sample(uniq_genre, 1)  
rec_genre <- sort(genre_t[as.character(s_genre), ], decreasing = TRUE)[1:n_recommended]
# get IDs
rec_id_genre <- as.numeric(names(rec_genre))
# create list 
rec_names <- anime_name[anime_name$anime_id %in% rec_id_genre, ]$name
# create table
header <- sprintf("Most viewed animes in %s", s_genre)
# display the list of similar animes
kable(rec_names, col.names = header) %>% kable_styling()
```
  
### Highly-Rated Animes With Low Popularity ("Surprise" Component)  
  
To add a surprise element to the list generated above, we will populate the top N recommendations by popularity using a list of animes that are the least popular but have high ratings.  
  
As highlighted in the chart from the [Relationship Between Number of Members and Ratings] section, there are some animes that might not be picked by our popularity system because those groups have few members. In fact, from that least popular group, only 2 animes are part of the popularity matrix calculated above. This can be seen below.  
  
```{r}
pop <- anime_comp %>% filter(log(members) > 3.5, log(members) < 4.5 , rating > 7 )
pop <- drop_na(pop)
tmp <- sum(c(as.character(pop$anime_id) %in% colnames(genre_t)))
tmp
```  
  
These two animes were dropped from this dataset and we built a "surprise"" popularity matrix using the same procedure as we'd used previously in the [Recommendations Based on Popularity] section.  

Then, a similarity matrix was interactively built based on the anime chosen in the [Top N Recommendations by Genre Similarity] section.  
  
```{r}
pop <- pop[!c(as.character(pop$anime_id) %in% colnames(genre_t)), ]
pop <- select(pop, c(1, 3, 6))
# splitting genre column
pop_g <- as.data.frame(tstrsplit(pop[, 2], '[,]', type.convert = FALSE), stringsAsFactors = FALSE)
colnames(pop_g) <- c(1:ncol(pop_g))
# grabing unique genres
g_vec <- (as.vector(trimws(unlist(pop_g))))
uniq_g <- unique(g_vec)
uniq_g <- uniq_g[!is.na(uniq_g)]
#creating matrix anime_id and genres
pop_genre_mtx <- cbind(anime_id = pop$anime_id, pop_g)
#creating binarized matrix anime_id and genre
pop_mtx <- matrix(NA, nrow(pop_genre_mtx), length(uniq_genre) + 1)
pop_mtx[, 1] <- pop$anime_id
colnames(pop_mtx) <- c("anime_id", uniq_genre)
for (i in 1:nrow(pop_genre_mtx)) {
 for (j in 2:ncol(pop_genre_mtx)) {
  mat_col = which(uniq_genre == trimws(pop_genre_mtx[i, j])) + 1
  pop_mtx[i, mat_col] <- 1
 }
}
genre_p <- pop_mtx
genre_p[, 2:ncol(pop_mtx)][is.na(genre_p[, 2:ncol(pop_mtx)])] <- 0
genre_p[, 2:ncol(pop_mtx)][genre_p[, 2:ncol(pop_mtx)] > 0] <- 1
genre_p <- genre_p[, 2:ncol(pop_mtx)]
# calculate similarity matrix - non normalized
tmp_anime <- genre_b[as.character(a_id), ]
genre_p <- rbind(genre_p, tmp_anime)
pop_bm <- as(genre_p, "binaryRatingMatrix")
pop_sim <- similarity(pop_bm, method = "Jaccard", which = "users")
# visualization - 
genre_pop <- as(pop_sim, "matrix")
colnames(genre_pop) <- c(pop$anime_id, a_id)
rownames(genre_pop) <- c(pop$anime_id, a_id)
image(genre_pop, main = "Similarity Matrix Between Least Popular Animes")
```

#### Top N Recommendations by Genre Similarity (with "Surprise")  
  
Below is the implementation of the recommendation list based on the genre similarity matrix we calculated above (which includes the "Surprise" kicker). One anime show is randomly sampled and then the N most similar animes to the one sampled are displayed.  
  
```{r genre rec}
# set number of recommendations
n_recommended <- 3
# get recommendations
rec_pop <- sort(genre_pop[as.character(a_id),], decreasing = TRUE)[1:n_recommended]
# get IDs
rec_id_p <- as.numeric(names(rec_pop))
# create list 
rec_names_p <- anime_name[anime_name$anime_id %in% rec_id_p,]$name
# create table
header <- sprintf("Surprise picks (lesser-known animes similar to): %s", a_name)
# display the list of similar animes
kable(rec_names_p, col.names = header) %>% kable_styling()
```  
  
## Collaborative Filtering Model  
  


# UBCF

```{r, ubcf, warning=FALSE}
# #ratings matrix
ubcf_rtg <- ratings_raw[ratings_raw$user_id %in% anime_genre_ids, ]
ubcf_rtg<- drop_na(ratings_raw)
ubcf_rtg <- as(ubcf_rtg, "realRatingMatrix")
ubcf_rtg<-ubcf_rtg[rowCounts(ubcf_rtg)>300,colCounts(ubcf_rtg)>1000] 
ubcf_mtx<-as(ubcf_rtg,"data.frame")
ubcf_rtg<-binarize(ubcf_rtg, minRating=1)

# # split datasets
eval_sets <- evaluationScheme(data = ubcf_rtg, method = "split", train=0.8, given = 1, goodRating = 1)
# # build UBCF model and SVD model
ubcf_rec <- Recommender(getData(eval_sets, "train"), "UBCF", param = list(method = "Jaccard"))
# # Make predictions with each model
ubcf_pred <- predict(ubcf_rec, getData(eval_sets, "known"), n = 10, goodRating = 1)
# Make predictions on whole dataset
ubcf_all<-predict(ubcf_rec,ubcf_rtg,n = 10, goodRating = 1)
# user sampling and recommendation display
id_sample<-sample(ubcf_mtx$user,1)
rec_ubcf<-recc_matrix[,as.character(id_sample)]
ubcf_animes <- anime_name[anime_name$anime_id %in% rec_ubcf,]$name
# create table
header <- sprintf("Top 10 recommendations for User: %s", id_sample)
# display the list of similar animes
kable(ubcf_animes, col.names = header) %>% kable_styling()
```  

## Display Recommendations in Shiny App
  
To allow us to generate predictions on the fly and display them, we created a Shiny App.

```{r shapp}
# library(shiny)
# runApp('App-1')
```

