---
title: "Final Project - DATA 612"
author: "Bruno de Melo and Leland Randles"
date: "July 16th, 2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(tidyverse)) install.packages("tidyverse",repos = "http://cran.us.r-project.org")
if(!require(recommenderlab)) install.packages("recommenderlab",repos = "http://cran.us.r-project.org")
if(!require(knitr)) install.packages("tidyverse",repos = "http://cran.us.r-project.org")
if(!require(kableExtra)) install.packages("kableExtra",repos = "http://cran.us.r-project.org")
if(!require(sparklyr)) install.packages("sparklyr",repos = "http://cran.us.r-project.org")
if(!require(NCmisc)) install.packages("NCmisc",repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table",repos = "http://cran.us.r-project.org")
library("tidyverse")
library("recommenderlab")
library("kableExtra")
library("sparklyr")
library("NCmisc")
library("data.table")
```  
  
## Loading Datasets

Per our final project proposal, there are two datasets we are using, which came from the web site https://myanimelist.net/, which is the world's largest anime and manga community.  

The datsets are available at kaggle: https://www.kaggle.com/CooperUnion/anime-recommendations-database.

### Content for Each Dataset  
  
* Anime.csv  
     
    + anime_id - myanimelist.net's unique id identifying an anime.     
    + name - full name of anime.    
    + genre - comma separated list of genres for this anime.     
    + type - movie, TV, OVA, etc.      
    + episodes - how many episodes in this show. (1 if movie).     
    + rating - average rating out of 10 for this anime.      
    + members - number of community members that are in this anime's "group".   
<br>
* Rating.csv              
     
    + user_id - non identifiable randomly generated user id.      
    + anime_id - the anime that this user has rated.     
    + rating - rating out of 10 this user has assigned (-1 if the user watched it but didn't assign a rating).  
  
We will use the ratings for collaborative filtering and genre plus type for content-based recommendation. Below, we load the datasets:   
  
```{r load, message = FALSE}
set.seed(1234)
# loading anime dataset
anime_raw <- read_csv("https://raw.githubusercontent.com/Randles-CUNY/DATA612/master/Final_Project/datasets_571_1094_anime.csv") %>% data.frame()  
anime_raw$name = str_replace_all(anime_raw$name,"&#039;","'")  
anime_raw$name = str_replace_all(anime_raw$name,"&quot;","")
anime_raw$name = str_replace_all(anime_raw$name,".hack//","") 
anime_raw$name = str_replace_all(anime_raw$name,"_","")
anime_raw$name = str_replace_all(anime_raw$name,"&amp;","&")
# useful data frames
anime_name <- select(anime_raw, c(1, 2))
anime_genre <- select(anime_raw, c(1, 3, 4))
anime_members <- select(anime_raw, c(1, 7))
anime_comp <- anime_raw
# loading ratings dataset
ratings_raw1 <- read_csv("https://raw.githubusercontent.com/bsvmelo/DATA612/master/Final_Project/rating_partaa.csv") %>% data.frame()
ratings_raw2 <- read_csv("https://raw.githubusercontent.com/bsvmelo/DATA612/master/Final_Project/rating_partab.csv", col_names = FALSE) %>% data.frame()
# converting to realRatingsMatrix
colnames(ratings_raw2) <- c(colnames(ratings_raw1))
ratings_raw <- rbind(ratings_raw1, ratings_raw2)
ratings_raw$rating[(ratings_raw$rating) == -1 ] <- NA
```  
  
## Data exploration  
  
### By type  
  
Based on the plot below, we decided exclude Music and NA from our recommender system dataset. We included TV, OVA (original video animation), Movie, Special and ONA (original net animation).  
  
```{r gg1, message = FALSE }
# Type histogram
a_type <- anime_comp %>% group_by(type) %>% summarise(Count=n())
ggplot(a_type, mapping = aes(x=reorder(type, -Count), y=Count)) + geom_col() + labs(x = "Anime Types")
```

### Relationship Between Number of Members and Ratings 
  
Per the data dictionary provided earlier, the `anime.csv` dataset includes two data elements of interest:  "members", which tells you how many community members joined the "group" associated with an anime show, and "rating" which is the average rating (out of 10) given by the anime group. Below, We check the relationship between the number of members and the ratings to check for biases.    
  
We used the log of the number of members in order to account for its variability (ranges from 5 to 1 million plus).  
  
The chart below indicates that there's a positive relationship between value of `members` and `rating`. We can also see in the rectangular area that there are less popular groups which have higher ratings, perhaps reflecting great anime shows which are more obscure. We may use this fact to enrich our recommendation list.   
  
```{r gg2, message=FALSE}
# size of communities vs rating plot
anime_comp <- drop_na(anime_comp)
ggplot(anime_comp, mapping = aes(x = log(members), y=rating)) + geom_point(alpha=1/15) + geom_smooth() + annotate("rect", xmin=c(3.3), xmax=c(4.7), ymin=c(7.3), ymax=c(10), alpha=0.2, color = "blue", fill = "orange")
```
  
## Content-Based Recommendation  
  
First, we built a content-based recommendation system based on genre similarity and popularity. The approach was to take the data element `genre` (which can contain multiple tagged genres) and build recommendations accordingly.  
  
### Recommendations Based on Genre Similarity  
  
First, we filtered the dataset by genre, getting rid of NAs values. Then, we create a binary matrix of animes by genre. Lastly, we calculate a similarity matrix.  
  
The utimate goal is to randomly select an anime, and display a top 10 recommendation list based on similarity between genres.  
  
```{r genre1}
# subsetting for type and NAs
anime_genre <- subset(anime_genre, type %in% c("TV", "OVA", "Movie", "Special", "ONA"), -type)
anime_genre <- drop_na(anime_genre) 
anime_genre_ids <- anime_genre$anime_id
# splitting genre column
genres1 <- as.data.frame(tstrsplit(anime_genre[,2], '[,]', type.convert = FALSE), stringsAsFactors = FALSE)
colnames(genres1) <- c(1:ncol(genres1))
# grabbing unique genres
genre_vec <- (as.vector(trimws(unlist(genres1))))
uniq_genre <- unique(genre_vec)
uniq_genre <- uniq_genre[!is.na(uniq_genre)]
# creating matrix anime_id and genres
anime_genre_mtx <- cbind(anime_id = anime_genre$anime_id, genres1)
# creating binarized matrix by anime_id and genre
genre_mtx <- matrix(NA, nrow(anime_genre), length(uniq_genre) + 1)
genre_mtx[,1] <- anime_genre_mtx$anime_id
colnames(genre_mtx) <- c("anime_id", uniq_genre)
for (i in 1:nrow(anime_genre_mtx)) {
 for (j in 2:ncol(anime_genre_mtx)) {
  mat_col = which(uniq_genre == trimws(anime_genre_mtx[i, j])) + 1
  genre_mtx[i, mat_col] <- 1
 }
}
# assigning binary values
genre_b <- genre_mtx
genre_b[, 2:ncol(genre_mtx)][is.na(genre_b[, 2:ncol(genre_mtx)])] <- 0
genre_b[, 2:ncol(genre_mtx)][genre_b[, 2:ncol(genre_mtx)] > 0] <- 1
genre_b <- genre_b[, 2:ncol(genre_mtx)]
# calculate similarity matrix - non-normalized
genre_bm <- as(genre_b, "binaryRatingMatrix")
genre_sim <- similarity(genre_bm, method = "Jaccard", which = "users")
# similarity matrix visualization - 20 animes
genre_v <- as(genre_sim, "matrix")
colnames(genre_v) <- anime_genre$anime_id
rownames(genre_v) <- anime_genre$anime_id
image(genre_v[1:20, 1:20], main = "Similarity Matrix Between 20 Animes")
```  
  
#### Top 10 Recommendations by Genre Similarity  
  
Below is the implementation of the recommendation list based on the similarity matrix we calculated above. One anime show is randomly sampled and then the 10 most similar animes to the one sampled are displayed.

```{r genrerec1}
# set number of recommendations
n_recommended <- 10
# randomly select an anime_id and get corresponding name 
a_id <- sample(anime_genre$anime_id, 1)
a_name <- anime_name[anime_name$anime_id == a_id, ]$name
# get recommendations
recs <- sort(genre_v[as.character(a_id),], decreasing = TRUE)[1:n_recommended]
# get IDs for recommended animes
recs_id <- as.numeric(names(recs))
# create list by name
recs_names <- anime_name[anime_name$anime_id %in% recs_id,]$name
# create table
header <- sprintf("Animes Similar to %s", a_name)
# display the table of similar animes
kable(recs_names, col.names = header) %>% kable_styling()
```
  
### Recommendations Based on Popularity   
  
We measured popularity by the number of times an anime has been viewed, based on the `ratings.csv` dataset. The ratings dataset is a typical user-item matrix. A frequency term was created using the function `colCounts` from the `recommenderLab` package.  
  
First, we conformed the rating dataset to the dataset used for Genre Similarity. Then, we created a viewer frequency or popularity dataset. We then transposed the binary matrix of animes by genres we'd created.  
  
In order to incorporate the popularity dataset into the binary matrix, our approach was to divide the number of views by the number of genres for the anime show. To avoid penalizing anime shows that have many genre classifications, the result was weighted by the square root of the total numbers of genres.  
  
The ultimate goal was to randomly select a genre and display a top 10 recommendation list based on the popularity index.  
  
```{r pop}
# subsetting the ratings matrix to match the user id used in the matrix above
rating_raw <- ratings_raw[ratings_raw$user_id %in% anime_genre_ids, ]
rating_mtx <- as(rating_raw, "binaryRatingMatrix")
user_mtx <- t(as(rating_mtx, "matrix"))
# counting how many times an anime has been seen
genre_count <- data.frame(anime_id = names(colCounts(rating_mtx)), views = colCounts(rating_mtx))
# creating anime-genre matrix 
rownames(genre_b) <- anime_genre$anime_id
genre_t <- t(genre_b)
colnames(genre_t) <- anime_genre$anime_id
genre_t <- genre_t[, colnames(genre_t) %in% c(as.character(genre_count$anime_id))]
#insert number of views in each column, weighted by square root of total numbers of genres
for (i in 1:ncol(genre_t)) {
  mat_col = which(colnames(genre_t)[i] == as.character(genre_count$anime_id))
  s <- genre_count$views[mat_col]/sqrt(sum(genre_t[, i]))
  genre_t[, i] <- genre_t[, i] * s
}
```
 
#### Top 10 Recommendations by Popularity
  
Below is the implementation of the recommender list based on the popularity matrix we'd calculated. To demonstrate, a genre is randomly sampled and the 10 most popular animes related to that genre are displayed.  
  
```{r genrerec2}
# set number of recommendations
n_recommended <- 10
# randomly select a anime_id and get corresponding name 
s_genre <- sample(uniq_genre, 1)  
rec_genre <- sort(genre_t[as.character(s_genre), ], decreasing = TRUE)[1:n_recommended]
# get IDs
rec_id_genre <- as.numeric(names(rec_genre))
# create list 
rec_names <- anime_name[anime_name$anime_id %in% rec_id_genre, ]$name
# create table 
header <- sprintf("Most viewed animes in %s", s_genre)
# display the list of similar animes 
kable(rec_names, col.names = header) %>% kable_styling()
```
  
### Highly-Rated Animes With Low Popularity ("Surprise" Component)  
  
To add a surprise element to the list generated above, we will populate the top 3 recommendations by popularity using a list of animes that are the least popular but have high ratings.  
  
As highlighted in the chart from the [Relationship Between Number of Members and Ratings] section, there are some animes that might not be picked by our popularity system because those groups have few members. In fact, from that least popular group, only 2 animes are part of the popularity matrix calculated above. This can be seen below.  
  
```{r sur1}
# identify overlap with previously created popularity matrix
pop <- anime_comp %>% filter(log(members) > 3.5, log(members) < 4.5 , rating > 7 )
pop <- drop_na(pop)
tmp <- sum(c(as.character(pop$anime_id) %in% colnames(genre_t)))
tmp
```  
  
These two animes were dropped from this dataset and we built a "surprise" popularity matrix using the same procedure as we'd used previously in the [Recommendations Based on Popularity] section.  

Then, a similarity matrix was interactively built based on the anime chosen in the [Top 10 Recommendations by Genre Similarity] section.  
  
```{r sur2}
# find animes in popularity dataset for all but the sample id
pop <- pop[!c(as.character(pop$anime_id) %in% colnames(genre_t)), ]
pop <- select(pop, c(1, 3, 6))
# splitting genre column
pop_g <- as.data.frame(tstrsplit(pop[, 2], '[,]', type.convert = FALSE), stringsAsFactors = FALSE)
colnames(pop_g) <- c(1:ncol(pop_g))
# grabing unique genres
g_vec <- (as.vector(trimws(unlist(pop_g))))
uniq_g <- unique(g_vec)
uniq_g <- uniq_g[!is.na(uniq_g)]
# creating matrix anime_id and genres
pop_genre_mtx <- cbind(anime_id = pop$anime_id, pop_g)
# creating binarized matrix anime_id and genre
pop_mtx <- matrix(NA, nrow(pop_genre_mtx), length(uniq_genre) + 1)
pop_mtx[, 1] <- pop$anime_id
colnames(pop_mtx) <- c("anime_id", uniq_genre)
for (i in 1:nrow(pop_genre_mtx)) {
 for (j in 2:ncol(pop_genre_mtx)) {
  mat_col = which(uniq_genre == trimws(pop_genre_mtx[i, j])) + 1
  pop_mtx[i, mat_col] <- 1
 }
}
# assigning binary values
genre_p <- pop_mtx
genre_p[, 2:ncol(pop_mtx)][is.na(genre_p[, 2:ncol(pop_mtx)])] <- 0
genre_p[, 2:ncol(pop_mtx)][genre_p[, 2:ncol(pop_mtx)] > 0] <- 1
genre_p <- genre_p[, 2:ncol(pop_mtx)]
# calculate similarity matrix - non-normalized
tmp_anime <- genre_b[as.character(a_id), ]
genre_p <- rbind(genre_p, tmp_anime)
pop_bm <- as(genre_p, "binaryRatingMatrix")
pop_sim <- similarity(pop_bm, method = "Jaccard", which = "users")
# similarity matrix visualization - 20 animes
genre_pop <- as(pop_sim, "matrix")
colnames(genre_pop) <- c(pop$anime_id, a_id)
rownames(genre_pop) <- c(pop$anime_id, a_id)
image(genre_pop, main = "Similarity Matrix Between 20 Animes")
```

#### Top 3 Recommendations by Genre Similarity (with "Surprise")  
  
Below is the implementation of the recommendation list based on the genre similarity matrix we calculated above (which includes the "Surprise" kicker). One anime show is randomly sampled and then the 3 most similar animes to the one sampled are displayed.  
  
```{r genrerec3}
# set number of recommendations
n_recommended <- 3
# get recommendations
rec_pop <- sort(genre_pop[as.character(a_id),], decreasing = TRUE)[1:n_recommended]
# get IDs
rec_id_p <- as.numeric(names(rec_pop))
# create list 
rec_names_p <- anime_name[anime_name$anime_id %in% rec_id_p,]$name
# create table
header <- sprintf("Surprise picks (lesser-known animes similar to): %s", a_name)
# display the list of similar animes
kable(rec_names_p, col.names = header) %>% kable_styling()
```  
  
## UBCF Model  
  
Finally, we built a user-based collaborative filtering recommender model based on data for users which had rated at least 300 anime and animes which had been rating by at least 1,000 users.  
  
```{r ubcf, warning=FALSE}
# create ratings matrix
ubcf_rtg <- ratings_raw[ratings_raw$user_id %in% anime_genre_ids, ]
ubcf_rtg <- drop_na(ratings_raw)
ubcf_rtg <- as(ubcf_rtg, "realRatingMatrix")
ubcf_rtg <- ubcf_rtg[rowCounts(ubcf_rtg) > 300, colCounts(ubcf_rtg) > 1000] 
ubcf_mtx <- as(ubcf_rtg, "data.frame")
ubcf_rtg <- binarize(ubcf_rtg, minRating = 1)
# split datasets
eval_sets <- evaluationScheme(data = ubcf_rtg, method = "split", train = 0.8, given = 1, goodRating = 1)
# build UBCF model
ubcf_rec <- Recommender(getData(eval_sets, "train"), "UBCF", param = list(method = "Jaccard"))
# make predictions
ubcf_pred <- predict(ubcf_rec, getData(eval_sets, "known"), n = 30, goodRating = 1)
# make predictions on whole dataset
ubcf_all <- predict(ubcf_rec, ubcf_rtg, n = 30, goodRating = 1) 
recc_matrix <- sapply(ubcf_all@items, function(x){colnames(ubcf_rtg)[x]})
# user sampling and recommendation display
id_sample <- sample(ubcf_mtx$user, 1)
rec_ubcf <- recc_matrix[ ,as.character(id_sample)]
ubcf_animes <- anime_name[anime_name$anime_id %in% rec_ubcf, ]$name
# create table
header <- sprintf("Top 10 recommendations for User: %s", id_sample)
# display the list of similar animes
kable(ubcf_animes[1:10], col.names = header) %>% kable_styling()
```  

## Display Recommendations in Shiny App
  
To allow us to generate and dispay predictions on the fly and display them (instead of just lists for one sample id, as we'd done in previous sections), we created a Shiny App:  

```{r sh, echo = FALSE}
shinyApp(
  # Define UI for dataset viewer app ----
ui <- fluidPage(
  
  # App title ----
  titlePanel("Anime Recommendations"),
  
  # Sidebar layout with a input and output definitions ----
  sidebarLayout(
    
    # Sidebar panel for inputs ----
    sidebarPanel(
      
      radioButtons("radio", h4("Desired Recommendation:"),
                   choices = c("Content-Based Genre" = "cbgenre", 
                               "Content-Based Anime" = "cbanime",
                               "Collaborative Filtering - Anime" = "cfanime")), 
      
      uiOutput("radiobuttonChoiceOutput"), 
      
      sliderInput("nbr_recs", h4("# of Anime to Recommend"),
                  min = 5, max = 30, value = 10)

      ),
    
    # Main panel for displaying outputs ----
    mainPanel(
      
      # Output: HTML table with requested number of observations ----
      tableOutput("reg_rec_list"), 
      tableOutput("sur_rec_list")
      
    )
  )
), 

# Define server logic to summarize and view selected dataset ----
server <- function(input, output) {

  output$radiobuttonChoiceOutput <- renderUI({
    switch(input$radio,
           cbgenre = selectInput("genre",
                       label = "Choose Genre",
                       choices = sort(uniq_genre)),  
           
           cbanime = selectInput("anime1",
                       label = "Choose Anime",
                       choices = sort(anime_name$name), 
                       selected = "Pokemon"), 
           
           cfanime = selectInput("users",
                       label = "Choose User",
                       choices = sort(unique(ubcf_mtx$user)), 
                       selected = 10072
                       )
                       
    )})
  
  # Show the first "n" observations ----
  output$reg_rec_list <- renderTable({
    if (input$radio == "cbgenre") {
    n_recommended <- input$nbr_recs
    s_genre <- input$genre
    rec_genre <- sort(genre_t[as.character(s_genre), ], decreasing = TRUE)[1:n_recommended]
    rec_id_genre <- as.numeric(names(rec_genre))
    rec_names <- anime_name[anime_name$anime_id %in% rec_id_genre, ]$name
    header <- sprintf("Most viewed Animes in %s", s_genre)
    tbl_reg <- data.frame(rec_names)
    names(tbl_reg)[names(tbl_reg) == "rec_names"] <- header
    }
    else if (input$radio == "cbanime") {
    n_recommended <- input$nbr_recs
    a_name = input$anime1
    a_id <- anime_name[anime_name$name == a_name, ]$anime_id
    recs <- sort(genre_v[as.character(a_id),], decreasing = TRUE)[1:n_recommended]
    recs_id <- as.numeric(names(recs))
    recs_names <- anime_name[anime_name$anime_id %in% recs_id, ]$name
    header <- sprintf("Animes Similar to %s", a_name)
    tbl_reg <- data.frame(recs_names)
    names(tbl_reg)[names(tbl_reg) == "recs_names"] <- header 
    }
    else {
    n_recommended <- input$nbr_recs
    u_id = input$users
    rec_ubcf <- recc_matrix[ ,as.character(u_id)]
    ubcf_animes <- anime_name[anime_name$anime_id %in% rec_ubcf, ]$name
    # create table
    header <- paste0(sprintf("Top %s", n_recommended),sprintf(" recommendations for User: %s", u_id))
    # display the list of similar animes
    tbl_reg <- data.frame(ubcf_animes[1:n_recommended])
    names(tbl_reg)[names(tbl_reg) == "ubcf_animes.1.n_recommended."] <- header 
    }
    tbl_reg
  })

  output$sur_rec_list <- renderTable({
    if (input$radio == "cbanime") {
    n_recommended <- 3
    a_name = input$anime1
    a_id <- anime_name[anime_name$name == a_name, ]$anime_id
    
    tmp_anime <- genre_b[as.character(a_id), ]
    genre_w <- rbind(genre_p[!rownames(genre_p) %in% "tmp_anime", ], tmp_anime)
    pop_bm <- as(genre_w, "binaryRatingMatrix")
    pop_sim <- similarity(pop_bm, method = "Jaccard", which = "users")
    genre_pop <- as(pop_sim, "matrix")
    colnames(genre_pop) <- c(pop$anime_id, a_id)
    rownames(genre_pop) <- c(pop$anime_id, a_id)
    
    n_recommended <- 3
    rec_pop <- sort(genre_pop[as.character(a_id), ], decreasing = TRUE)[1:n_recommended]
    rec_id_p <- as.numeric(names(rec_pop))
    rec_names_p <- anime_name[anime_name$anime_id %in% rec_id_p,]$name
    header <- sprintf("Surprise Picks (Lesser-Known Animes Similar to): %s", a_name)
    tbl_sur <- data.frame(rec_names_p)
    names(tbl_sur)[names(tbl_sur) == "rec_names_p"] <- header 
    genre_w <- NA
    }
    else {
      tbl_sur <- NULL
    }
    tbl_sur
  })  

}, 
options = list(height = 1500)
)
```

